Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-19 index.qmd
section:preliminaries,theory; process:write,images

@@

@@

@@

@@

@@

## Why estimate judges appart? {#sec-appA}

>>>>

Causal analysis is the field of experimental design and statistics pertaining to establishing cause and effect from data.

Causal inference

Directed Acyclic Graph (DAG)

the four components of DAGs

Structural Causal Model (SCM)

the four components of DAGs in SCM notation

@@

![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=100%}

@@

![ Identification-Estimation flowchart. Extracted from @Neal_2020 [pp. 32] ](images/figures/IEflow.png){#fig-IEflow fig-align="center" width=40%}

@@

![ DCJ causal diagram, simplified description ](images/figures/SciModel_simp1.png){#fig-SciModel_simp1 fig-align="center" width=80%}

![ DCJ causal diagram, simplified mathematical description ](images/figures/SciModel_simp2.png){#fig-SciModel_simp2 fig-align="center" width=70%}

![ DCJ causal diagram, population mathematical description ](images/figures/SciModel_pop.png){#fig-SciModel_pop fig-align="center" width=70%}

<!-- check the directionality of the arrow from $p_{I}$ to $\sigma_{i(g)}$ and equivalents -->

![ DCJ causal diagram, sample with comparisons mathematical description ](images/figures/SciModel_samp.png){#fig-SciModel_samp fig-align="center" width=80%}

## Probabilitics assumptions of the scientific model {#sec-theory-probability}

<!-- what does it mean in probabilistic terms? describe probabilities and arrive to structural model -->

<!-- commands for d-separation -->
\newcommand{\dsep}{\perp\!\!\!\perp}
\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}

$$
\begin{aligned}
  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
  \delta_{kijr_{ij}} := & \; f_{D}( \gamma_{k}, \theta_{ir_{i}} ) \\
  \gamma_{k} := & \; f_{G}( Z_{k}, e_{k} ) \\
  \theta_{ir_{i}} := & \; f_{R}( \theta_{i}, Y_{r}, e_{r_{i}} ) \\
  \theta_{i} := & \; f_{T}( X_{i}, e_{i} ) \\
  & \; e_{k} \dsep e_{i} \\
  & \; e_{k} \dsep e_{r_{i}} \\
  & \; e_{i} \dsep e_{r_{i}}
\end{aligned}
$$ {#eq-StructuralModel}

@@

$$
\begin{aligned}
  O_{kijr_{ij}} \sim & \; \text{Bernoulli}\left[ \; logit^{-1}\left( \delta_{kijr_{ij}} \right) \; \right] \\
  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
  \gamma_{k} = & \; logit^{-1}\left[ \beta_{Z} Z_{k} + e_{k} \right] \\
  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
\end{aligned}
$$ {#eq-StatModel_general}

for identification purposes we can set $\frac{1}{G}\sum_{g=1}^{G} \sigma_{k(g)} = 0.02$, $\frac{1}{G}\sum_{g=1}^{G} \sigma_{i(g)} = 1$, and $\frac{1}{G}\sum_{g=1}^{G} \sigma_{r(g)} = 1$. A special case of this would be to assume that the data comes from the same population, in that case, $\sigma_{k(g)} = \sigma_{k} = 0.02$, $\sigma_{i(g)} = \sigma_{i} = 1$

## Let's talk about Thurstone {#sec-theory-thurstone}

Thurstone's comparative judgment @Thurstone_1927 is based on the formula:
$$
X_{AB} = \frac{S_{A} - S_{B}}{\sigma_{AB}}
$$

where $X_{AB}$ defines the comparative judgment outcome, $S_{A}$ and $S_{B}$ are the modal discriminal processes, $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B} + 2 \rho \sigma_{A} \sigma_{B}}$, with $\sigma_{A}$ and $\sigma_{B}$ being the dispersion of discriminal processes $A$ and $B$, respectively, and $\rho$ the correlation between discriminal processes.

The theory identifies five cases:

- **Case 1:** only constant $\rho$ (not $\rho_{ij}$)
- **Case 2:** $X_{ij}$ becomes $X_{kij}$ with $k=1, \dots, K$ judges (replication, not duplication)
- **Case 3:** $\rho = 0$, then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$ 
- **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$
- **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$

Now using the DAG and statistical notation

$$
\begin{aligned}
  O_{kijr_{ij}} := & \; f_{O}( \delta_{kijr_{ij}} ) \\
  \delta_{kijr_{ij}} = & \; \gamma_{k}( \theta_{ir_{i}} - \theta_{jr_{j}} ) \\
  \gamma_{k} = & \; f_{G}( Z_{k}, e_{k} ) \\
  \theta_{ir_{i}} = & \; \theta_{i} + \beta_{Y} Y_{r} + e_{r_{i}}  \\
  \theta_{i} = & \; \beta_{X} X_{i} + e_{i}  \\
  e_{k} \sim & \; \text{Normal}(0, \sigma_{k(g)}) \\
  e_{i} \sim & \; \text{Normal}(0, \sigma_{i(g)}) \\
  e_{r_{i}} \sim & \; \text{Normal}(0, \sigma_{r(g)})
\end{aligned}
$$ {#eq-StatModel_thurstone}

The theory identifies five cases:

- **Case 1:** only constant $\rho \approx \sigma_{i}$
- **Case 2:** now judges are separated by using $\gamma_{k}$
- **Case 3:** $\rho \approx \sigma_{e_{i}} = 0$ (no nesting of texts on students), then $\sigma_{AB} = \sqrt{ \sigma^{2}_{A} + \sigma^{2}_{B}}$ 
- **Case 4:** $\sigma_{B}=\sigma_{A}+d$, then $\lim_{d \leq 0.1\sigma_{A}} \sigma_{AB} = (\sigma_{A} + \sigma_{B}) /\sqrt{2}$
- **Case 5:** $\sigma_{B}=\sigma_{A}$, then $\sigma_{AB} = \sqrt{2}\sigma$

But now can we see other scenarios than just those 5 cases?

- consider different $\rho \approx \sum_{p=1}^{P} \sigma_{p}$, depending on $P$ nesting structures 
- we can now investigate $\gamma_{k}$
- we can assume $\sigma_{B} \neq \sigma_{A}$, no need for results on the limit

@@

## Why do we need to estimate judges' abilities? {#sec-appA}
